---
title: "p8105_hw2_mc5503.Rmd"
author: "mc5503"
date: '2023-10-01'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(readxl)
```
# Problem 1
## tidy pols data
```{r}
pols=read_csv("data/fivethirtyeight_datasets/pols-month.csv")|>
  separate(mon,into = c("year","month_num","day"),convert = TRUE)|>
  mutate(month=month.name[month_num])|>
  mutate(president=recode(prez_gop,"0"="dem","1"="gop","2"="gop"))|>
  select(year,month,everything(),-day,-prez_gop,-prez_dem,-month_num)
```
## tidy snp data
```{r}
snp=read_csv("data/fivethirtyeight_datasets/snp.csv") |>
  separate(date, into = c("month", "day", "year"), convert = TRUE) |>
  mutate(year=ifelse(year>=50,1900+year,2000+year))|>
  arrange(year, month) |>
  mutate(month = month.name[month]) |>
  select(year, month, close) 
```
## tidy unemployment data
```{r}
month_df = 
  tibble(
    month_num = 1:12,
    month_abbr = month.abb,
    month = month.name
  )

unemployment = 
  read_csv("data/fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec,
    names_to="month_abbr",
    values_to ="unemployment"
  )|>
  left_join(x=_,y=month_df)|>
  select(year,month,unemployment)
```
## merge all the datasets
```{r}
data_538=
  left_join(pols,snp)|>
  left_join(x=_,y=unemployment)

data_538
str(data_538)
```
The `pols` dataset has `r nrow(pols)` observations and `r ncol(pols)` variables. It contains information about the party affiliation distribution from years `r range(pols$year)[1]` to `r range(pols$year)[2]`.
The `snp` data has `r nrow(snp)` observations and `r ncol(snp)` variables, ranging from years `r range(snp$year)[1]` to `r range(snp$year)[2]`.
The `unemployment` data has `r nrow(unemployment)` observations and `r ncol(unemployment)` variables ranging from years `r range(unemployment$year)[1]` to `r range(unemployment$year)[2]`.
When a democrat was president, the average unemployment rate was `r filter(data_538, month == "January", year >= 1975, president == "dem") |> pull(unemployment) |> mean() |> round(2)`. While for republician, the average unemployment rate was `r filter(data_538, month == "January", year >= 1975, president == "gop") |> pull(unemployment) |> mean() |> round(2)`.

# Problem 2
## import and clean the datasets
```{r}
mr_tw=read_excel(
  "data/202309 Trash Wheel Collection Data.xlsx",
   sheet = "Mr. Trash Wheel",
  .name_repair = "unique_quiet")|>
  janitor::clean_names()|>
  select(-starts_with("x"))|>
  filter(dumpster!="NA")|>
  mutate(homes_powered=round(weight_tons*500/30))|>
  mutate(trash_wheel_name="Mr. Trash Wheel")|>
  mutate(year=as.numeric(year))
  
pro_tw=read_excel(
  "data/202309 Trash Wheel Collection Data.xlsx",
  sheet = "Professor Trash Wheel")|>
  janitor::clean_names()|>
  select(-starts_with("x"))|>
  filter(dumpster!="NA")|>
  mutate(homes_powered=round(weight_tons*500/30))|>
  mutate(trash_wheel_name="Professor Trash Wheel")

gwy_tw=read_excel(
  "data/202309 Trash Wheel Collection Data.xlsx",
  sheet = "Gwynnda Trash Wheel")|>
  janitor::clean_names()|>
  select(-starts_with("x"))|>
  filter(dumpster!="NA")|>
  mutate(homes_powered=round(weight_tons*500/30))|>
  mutate(trash_wheel_name="Gwynnda Trash Wheel")
```

## combine all the datasets
```{r}

tw_tidy=bind_rows(mr_tw,pro_tw,gwy_tw)|>
  select(date,trash_wheel_name,everything())

str(tw_tidy)
```
The `mr_tw`, `pro_tw` and `gwy_tw` datasets tells about the weight of trash, trash types and the quantity of each type.There were `r max(mr_tw$dumpster)` dumpsters used by Mr. Trash Wheel ranged from `r range(mr_tw$date)[1]` to `r range(mr_tw$date)[2]`. The start date `r range(mr_tw$date)[1]` might be wrong and it might be `r mr_tw$date[1]`.
There are `r max(pro_tw$dumpster)` dumpsters used by Professor Trash Wheel ranged from `r range(pro_tw$date)[1]` to `r range(pro_tw$date)[2]`. There were `r max(gwy_tw$dumpster)` dumpsters used by Gwynnda Trash Wheel ranged from `r range(gwy_tw$date)[1]` to `r range(gwy_tw$date)[2]`.

The `tw_tidy` dataset contains information of all the dumspters used by the three wheels.

Total weight of trash collected by Professor Trash Wheel is `r sum(pro_tw$weight_tons)` tons. The total number of cigarette butts collected by Gwynnda in July of 2021 is `r filter(gwy_tw, month %in% c("july","July"),year == 2021) |> pull(cigarette_butts) |> sum()`.

# Problem 3
## import and tidy the dataset of baseline demographics
```{r}
mci_baseline=read_csv("data/data_mci/MCI_baseline.csv",skip = 1)|>
  janitor::clean_names()|>
  mutate(sex=recode(sex,"0"="female","1"="male"))|>
  mutate(apoe4=recode(apoe4,"1"="carrier","0"="non-carrier"))|>
  filter(age_at_onset>current_age|age_at_onset==".")
```
Using `mutate` to replace 0 and 1 to character variable, then using `filter` to delete the participant who develop mci at or before the baseline.

There are total `r max(mci_baseline$id)` participants in the study and `r nrow(mci_baseline)` participants were eligible. `r sum(mci_baseline$age_at_onset>0)` participants developed MCI. The average baseline age is `r mean(mci_baseline$current_age)`. The proportion of women in the study who are APOE4 carriers is `r nrow(filter(mci_baseline,sex=="female",apoe4=="carrier"))/nrow(filter(mci_baseline,sex=="female"))`.

## import and tidy the dataset of longitudinally observed biomarker values
```{r}
mci_amyloid=read_csv("data/data_mci/mci_amyloid.csv",skip=1)|>
  janitor::clean_names()|>
  rename("id"="study_id")
```
```{r}
mci_amyloid_pivot=pivot_longer(
  mci_amyloid,
  time_2:time_8,
  names_to = "time",
  values_to = "amyloid_ratio")|>
  drop_na()|>
  mutate(time=recode(time,"time_2"="2 years","time_4"="4 years","time_6"="6 years","time_8"="8 years"))
```

Use `rename` to change study_id to id, which maps the `mci_baseline` data. Use `pivot_longer` to make the data more tidy and `drop_na` to delete the missing values in amyloid ratio.There are total `r nrow(mci_amyloid)` participants and `r nrow(mci_amyloid_pivot)` records about amyloid ratio in the study .

## combine two datasets
```{r}
anti_join(mci_baseline,mci_amyloid,by="id")
anti_join(mci_amyloid,mci_baseline,by="id")
mci_data=inner_join(mci_baseline,mci_amyloid,by="id")
str(mci_data)
```
There are `r nrow(anti_join(mci_baseline,mci_amyloid,by="id"))` participants appear only in the baseline dataset (after deleting the ineligible participant) and `r nrow(anti_join(mci_amyloid,mci_baseline,by="id"))` participants (including the ineligible people in the baseline dataset) appear only in amyloid dataset.

After combining two datasets, there are `r nrow(mci_data)` participants in both study, including `r nrow(filter(mci_data,apoe4=="carrier"))` apoe4 carriers. There are `r nrow(filter(mci_data,sex=="female"))` female in total.

## export the mci_data
```{r}
write_csv(mci_data,"data/data_mci/mci_data.csv")
```

