---
title: "p8105_hw2_mc5503.Rmd"
author: "mc5503"
date: '2023-10-01'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(readxl)
```
# Problem 1
## tidy pols data
```{r}
pols=read_csv("data/fivethirtyeight_datasets/pols-month.csv")|>
  separate(mon,into = c("year","month_num","day"),convert = TRUE)|>
  mutate(month=month.name[month_num])|>
  mutate(president=recode(prez_gop,"0"="dem","1"="gop"))|>
  select(year,month,everything(),-day,-prez_gop,-prez_dem,-month_num)
```
## tidy snp data
```{r}
snp=read_csv("data/fivethirtyeight_datasets/snp.csv") |>
  separate(date, into = c("month", "day", "year"), convert = TRUE) |>
  mutate(year=ifelse(year>=50,1900+year,2000+year))|>
  arrange(year, month) |>
  mutate(month = month.name[month]) |>
  select(year, month, close) 
```
## tidy unemployment data
```{r}
month_df = 
  tibble(
    month_num = 1:12,
    month_abbr = month.abb,
    month = month.name
  )

unemployment = 
  read_csv("data/fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec,
    names_to="month_abbr",
    values_to ="unemployment"
  )|>
  left_join(x=_,y=month_df)|>
  select(year,month,unemployment)
```
## merge all the datasets
```{r}
data_538=
  left_join(pols,snp)|>
  left_join(x=_,y=unemployment)

data_538
str(data_538)
```
The `pols` dataset has `r nrow(pols)` observations and `r ncol(pols)` variables. It contains information about the party affiliation distribution from years `r range(pols$year)[1]` to `r range(pols$year)[2]`.
The `snp` data has `r nrow(snp)` observations and `r ncol(snp)` variables, ranging from years `r range(snp$year)[1]` to `r range(snp$year)[2]`.
The `unemployment` data has `r nrow(unemployment)` observations and `r ncol(unemployment)` variables ranging from years `r range(unemployment$year)[1]` to `r range(unemployment$year)[2]`.
When a democrat was president, the average unemployment rate was `r filter(data_538, month == "January", year >= 1975, president == "dem") |> pull(unemployment) |> mean() |> round(2)`. While for republician, the average unemployment rate was `r filter(data_538, month == "January", year >= 1975, president == "gop") |> pull(unemployment) |> mean() |> round(2)`.
